---
title: "Pathway Analysis"
author: "by Alexander Pico"
package: rWikiPathways
date: "`r Sys.Date()`"
output: 
  BiocStyle::html_document:
    toc_float: true
#  pdf_document:
#    toc: true 
vignette: >
  %\VignetteIndexEntry{4. Pathway Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r, echo = FALSE}
knitr::opts_chunk$set(
  eval=FALSE
)
```

This vignette will cover a wide range of analytical and visualization techniques involved in a typical pathway analysis. The *Overview* section will go into more detail on the particulars, but note that this vignette is designed to be modular and carefully considered. Please do not simply run the entire script and expect to get any meaningful from the final output. This is an instructional device, ideal for guided workshops.

As you migth expect, a broad tour of pathway anlaysis will require a variety of *R* packages. Please make sure you have the latest *R* installed (3.5 or later) and review the output from running the next code chunck. The final printed line will state whether it was successful or not. Good luck!

# Installation and Setup
```{r}
load.libs <- c(
  "DOSE",
  "GO.db",
  "GSEABase",
  "org.Hs.eg.db",
  "clusterProfiler",
  "dplyr",
  "tidyr",
  "ggplot2",
  "stringr",
  "RColorBrewer",
  "rWikiPathways",
  "RCy3")

install.libs <- load.libs[!load.libs %in% installed.packages()]
source("https://bioconductor.org/biocLite.R")
biocLite(install.libs)
status <- sapply(load.libs,require,character.only = TRUE)
if(all(status)){
    print("Successfully installed and loaded all required libraries.")
} else{
    cat("One or more libraries failed to install correctly. Check the following list for FALSE cases and try again...\n\n")
    status
}
```

The RCy3 package is used to connect with Cytoscape. So you will also need to install and launch Cytoscape:

* Download the latest Cytoscape from http://www.cytoscape.org/download.php
* Complete installation wizard
* Launch Cytoscape

```{r}
cytoscapePing()  #this will tell you if you're able to successfully connect to Cytoscape or not
```

For this vignette, you'll also need a couple apps for Cytoscape. With Cytoscape running, you can install each of these from the Cytoscape App Store with a single click:
* http://apps.cytoscape.org/apps/wikipathways
* http://apps.cytoscape.org/apps/cytargetlinker 

If you are running Cytoscape 3.7.0 or above, you can simply run this command:
```
installApp('WikiPathways') 
installApp('CyTargetLinker') 
```

# Overview 


# Dataset
```{r}
lung.expr <- read.csv(system.file("extdata","data-lung-cancer.csv", package="rWikiPathways"),stringsAsFactors = F)
nrow(lung.expr)
head(lung.expr)

up.genes <- lung.expr[lung.expr$log2FC > 1 & lung.expr$adj.P.Value < 0.05, 1]
dn.genes <- lung.expr[lung.expr$log2FC < 1 & lung.expr$adj.P.Value < 0.05, 1]
bkgd.genes <- lung.expr[,1]
```

# Enrichment
```{r}
up.genes.entrez <- clusterProfiler::bitr(up.genes,fromType = "ENSEMBL",toType = "ENTREZID",OrgDb = org.Hs.eg.db)
cat("\n\nWhich column contains my new Entrez IDs?\n")
head(up.genes.entrez[[1]])
head(up.genes.entrez[[2]])
```

```{r}
# supported sources of identifiers (and precise spelling)...
keytypes(org.Hs.eg.db)
```

```{r}
dn.genes.entrez <- bitr(dn.genes,fromType = "ENSEMBL",toType = "ENTREZID",OrgDb = org.Hs.eg.db)
bkgd.genes.entrez <- bitr(bkgd.genes,fromType = "ENSEMBL",toType = "ENTREZID",OrgDb = org.Hs.eg.db)
```

## Gene Ontology
```{r}
egobp <- clusterProfiler::enrichGO(
        gene     = up.genes.entrez[[2]],
        universe = bkgd.genes.entrez[[2]],
        OrgDb    = org.Hs.eg.db,
        ont      = "BP",
        pAdjustMethod = "fdr",
        pvalueCutoff = 0.05, #p.adjust cutoff (https://github.com/GuangchuangYu/clusterProfiler/issues/104)
        readable = TRUE)

head(egobp,10)
```

A variety of plots are provided by *clusterProfilier*...
```{r}
barplot(egobp, showCategory = 20)
dotplot(egobp, showCategory = 20)
emapplot(egobp, showCategory = 20)
goplot(egobp)
```

For more control and customization, here is an example using **ggplot** and the same **enrichGO object** output from *clusterProfilier*...
```{r}
ggplot(egobp[1:20], aes(x=reorder(Description, -pvalue), y=Count, fill=-p.adjust)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    scale_fill_continuous(low="blue", high="red") +
    labs(x = "", y = "", fill = "p.adjust") +
    theme(axis.text=element_text(size=11))
```
Use **str(egobp)** to explore the data structure in more detail to see what you can use in plotting.
 
## WikiPathways
That's all well and fine, but we are here for pathway analysis!  Everything up to this point is basic GO analysis. Building on top of this foundation, let's see what WikiPathways can add. The *clusterProfiler* package doesn't include built-in support for WikiPathways (yet), but it does support the GMT standard file format and so does WikiPathways! It just takes a few commands to prepare WikiPathways data for clusterProfiler's **enricher** function, which is the generic analog of **enrichGO**.

The first step is to retreive the latest GMT file listing all the genes in each of the human pathways. WikiPathways releases a GMT file per species every month and archives them for reproducible research. Here, we'll use the first function from the *rWikiPathways* package, aptly named **downloadPathwayArchive**...

```{r}
wp.hs.gmt <- rWikiPathways::downloadPathwayArchive(organism="Homo sapiens", format = "gmt")

# supported organisms (and case-dependent spelling)...
listOrganisms()
```

Now that we've got the latest GMT file for human pathways, we can feed it into *clusterProfiler* and generate the two dataframes we need for **enricher**...
```{r}
wp2gene <- clusterProfiler::read.gmt(wp.hs.gmt)
wp2gene <- wp2gene %>% separate(ont, c("name","version","wpid","org"), "%")
wpid2gene <- wp2gene %>% select(wpid,gene) #TERM2GENE
wpid2name <- wp2gene %>% select(wpid,name) #TERM2NAME
```

That's it! We are ready to perform pathway analysis...
```{r}
ewp.up <- clusterProfiler::enricher(
        up.genes.entrez[[2]],
        universe = bkgd.genes.entrez[[2]],
        pAdjustMethod = "fdr",
        pvalueCutoff = 0.1, #p.adjust cutoff
        TERM2GENE = wpid2gene,
        TERM2NAME = wpid2name)

head(ewp.up)
```

For some reason, **enricher** doesn't automatically add gene symbols to the result object, but there is a handy function in **DOSE** that does...
```{r}
ewp.up <- DOSE::setReadable(ewp.up, org.Hs.eg.db, keytype = "ENTREZID")
head(ewp.up)
```

And, we have access to all the same plotting functions as before...
```{r}
barplot(ewp.up, showCategory = 20)
dotplot(ewp.up, showCategory = 20)
emapplot(ewp.up, showCategory = 20)
```

Before we forget, we can also do the same analysis for down-regulated gene...
```{r}
ewp.dn <- enricher(
  dn.genes.entrez[[2]],
  #universe = bkgd.genes[[2]],  #hint: comment out to get _any_ results
  pAdjustMethod = "fdr",
  pvalueCutoff = 0.1, #p.adjust cutoff
  TERM2GENE = wpid2gene,
  TERM2NAME = wpid2name)

 ewp.dn <- setReadable(ewp.dn, org.Hs.eg.db, keytype = "ENTREZID")
 head(ewp.dn)
 dotplot(ewp.dn, showCategory = 20)
```
Interesting... Almost twice as many down-regulated genes (634 vs 366), but fewer significant pathway hits. Pathway analysis is a more focused approach than GO analysis. It requires sets of genes that are *functionall* related in the context of known pathways mechanisms.  So, gene set size doesn't always correlate with result size.

Ok. These built-in plots are good for displaying the basic results, but one of the advantages of pathway analysis is that you have *pathway models* already built and ready for data overlays. We will get to that soon in the in *Envision* section, but first, let's see what else we can learn about the pathways hits from WikiPathways.

# Explore
Turning again to the *rWikiPathways* package, let's explore the content and some of our pathway hits so far. Since we are studying Lung Cancer here, let's start with a search for relevant pathways...
```{r}
findPathwayNamesByText("lung cancer")
```

Whoa, that's a lot... and there's a bunch of repeats!?  This general search includes all matches to "lung" and/or "cancer", sorted by best matches to both terms. And it also includes matches from all speceis, e.g., mouse and rat, in addition to human.  Let's be more specific...

```{r}
lc.pathways <- findPathwaysByText('"lung cancer"')  #quotes inside query to require both terms
human.filter <- lapply(lc.pathways, function(x) x$species == "Homo sapiens")
human.lc.pathways <- lc.pathways[unlist(human.filter)] # just the human lung cancer pathways
sapply(human.lc.pathways, function(x) x$name) # display the pathway titles
```

Ok, so there are 3 pathways that explicitly mention "lung cancer" in their titles or descriptions. None of these were in our top hits for enrichment, but we might want to look at them anyways during our exploratory data visualization, right?  So, let hold on to their WPIDs for now...

```{r}
lc.wpids <- sapply(human.lc.pathways, function(x) x$id)
lc.wpids
```

You can also search pathways by gene identifiers, pubmed references and ontology terms.  But we already know the primary pathways we want to see based on our pathway enrichment analysis. Let's identify their WPIDs...

```{r}
ewp.up.wpids <- ewp.up$ID
ewp.up.wpids
```

Let's take a look at these. We could open them in our browser or we could download SVG image files... Let's do both!
```{r}
url <- getPathwayInfo("WP179")[2]
browseURL(url)

svg <- getColoredPathway('WP179')
writeLines(svg, 'WP179.svg')
```

You can access all of the WikiPathways website information using *rWikiPathways*. You can even query the history of a particular pathway or recent changes across the entire site. It's a wiki afterall!

But what we really want to do next is view our data on these pathways. For that, we are going to turn to *Cytoscape* and the *RCy3* package.

# Envision
*Cytoscape* is a popular network visualization and analysis tool with great community support for development and scripting. Since pathways are just a special type of network, it's perfect for providing high quality visualization for pathway analysis results.

We have the *RCy3* package loaded already, but we also have to have *Cytoscape* launched (see Step 1. Installation, if you haven't already). Once *Cytoscape* is running, try to **ping** it with this command:

```{r}
cytoscapePing()
```

If you've got everything loaded and running, then all you need to do is run this command to import a pathway into Cytoscape from WikiPathways:
```{r}
commandsRun('wikipathways import-as-pathway id=WP179') 
```

load data
set style

import all other pathways
load data
apply style

# Extend
cytargetlinker

# Export
```{r}
save(ewp.up, file = "lung_cancer_ewp_up.Rdata")
save(ewp.dn, file = "lung_cancer_ewp_down.Rdata")
```


And with so many libraries involved, it's a good idea to keep track of the version information when performing real analyses so that you (and others) can reliably reproduce your results:
```{r}
sessionInfo()
```